Fast Sorting on Modern Computers

### 2005-05-30 paper/docs/heap.tex
finished heap section

The most interesting thing is the relation between the predictablility of the
comparisons in the destruction of the 2-, 4- and 8-heaps, which are shown in
Figure \ref{Branch prediction performance for base and cache heapsorts}.
Overall, the 4-heap is half as high as the 2-heap, and twice as high as the
8-heap, but has to make more comparisons. However, there are three comparisons
in the 4-heap and seven comparisons in the 8-heap. In total, there are 50\% more
comparisons in the 4-heap, and 150\% more in the 8-heap.

These branches become more predictable. The first predicts correctly 50\% of the
time, while the next predicts correctly 60\% of the time, then 71\%, 77\%, and
so on in a curve. This curve is shown in Figure TODO.
%TODO make a pretty graph for this.

Also notable from the results is that branches are more often not-taken then
they are correctly predicted, implying that a semi-static predictor based on
hints from the compiler would do a better job at making correct predictions.

>>>>

The most interesting thing is the relation between the predictablility of the
comparisons during the destruction of the 2-, 4- and 8-heaps. Compare the
columns marked `\cc{Destruction, child comparison}' and `\cc{Destruction, child
promotion}' in Figure \ref{base and cache heapsorts-1} (the 2-heap), with the
columns marked `\cc{comparison 1}' to `\cc{child promotion}' in Figure \ref{base and
cache heapsorts-2} (the 4-heap) and Figures \ref{base and cache heapsorts-3} and
\ref{base and cache heapsorts-4} (the 8-heap).

Though the number of times each comparison is used shrinks
as the size of the fanout increased, overall the number of branches increases.
In total, there are 50\% more comparisons in the 4-heap than the 2-heap, and
150\% more in the 8-heap.

More importantly, the accuracy of the each
successive comparison increases in the same way as selection sort (see Section
\ref{selection branches logn}), becoming more predictable as the fanout
increases. For example, in the 8-heap, shown in Figures \ref{base and cache heapsorts-3}
and \ref{base and cache heapsorts-4},
the first comparison is predicted correctly 50\% of the
time, while the next is predicted correctly 61\% of the time, then 71\%, 77\%, and
so on in a curve, which is shown in Figure \ref{cache heapsort curve}.

\begin{figure}
\includegraphics{../counters/curve.eps}
\caption{Predictability of branches in cache heapsort, with a fanout of eight}
\label{cache heapsort curve}
\end{figure}

\label{heapsort entropy}
We observe here that the more branches there are, the more are correctly
predicted, and that there is a threshold below which the number of branches will
not drop. This is discussed in more Section \ref{quicksort entropy}.

Also notable from the results is that branches are slightly more often not-taken
then they are correctly predicted, implying that a semi-static predictor based
on hints from the compiler could do a better job at making correct predictions.
